Supplementary Materials: Separating random and deterministic sources of computational noises in explore-exploit decisions

### 2025-04-19 Plos Comp Bio - revision v1/manuscript_DeterministicRandomNoise.tex
u

% control the fontsize
\PassOptionsToPackage{textsize=scriptsize}{todonotes}
\usepackage[]{changes}
\usepackage{changes}
\definechangesauthor[name={Siyu Wang}, color=red]{siyu}
\definechangesauthor[name={Bob}, color=blue]{bob}

@@

Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (see Supplementary Materials section 2.2). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. 
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

@@

\subsubsection*{Model-based results} 
Posterior distributions over the group-level means of the deterministic and random noise standard deviation $\sigma_{det}$ and $\sigma_{ran}$ are shown in Figure \ref{fig:mb1} and Supplementary Figure S11. Consistent with our model-free results, we see that both random and deterministic noise are non-zero. Numerically, random noise is about 2-3 times larger than the deterministic noise. By computing the posterior distribution of $\sigma^2_{det}/(\sigma^2_{det}+\sigma^2_{ran})$, our data suggests that 14.25\% of the variability in random exploration is accounted for by deterministic noise ([4.90\%, 28.81\%], 95\% CI). In addition, we find that both random and deterministic noise increase with horizon. This increase was larger for random noise (mean = 7.13, 100\% of samples showed an increase in random noise with horizon) than deterministic noise (mean = 2.59, 98.64\% of samples showed an increase in deterministic noise with horizon). But intriguingly, the relative increase in both types of noise was similar (Figure \ref{fig:ratio}). That is, when we compute the relative increase in deterministic noise with horizon, $\sigma^{det}_{horizon6}/\sigma^{det}_{horizon1}$, it is very similar to the relative increase in random noise with horizon $\sigma^{ran}_{horizon6}/\sigma^{ran}_{horizon1}$.

@@

To perform this `posterior predictive check', we created a set of simulated data by taking the subject-level parameters from the hierarchical Bayesian fits and having the model play the same sequence of games as seen by the subjects. We then applied the same model-free analysis as described in the previous sections to this simulated data set and compared the model's behavior to that of participants. As shown in Figure \ref{fig:mb3}, the model can account for all qualitative patterns in the data --- the increase in p(high info), p(low mean), and p(inconsistent) with horizon, and that p(inconsistent) is in between pure random and pure deterministic noise. The quantitative agreement is almost perfect for p(high info) and for p(inconsistent) in the [1 3] condition, but the model seems to systematically overestimate p(low mean) and p(inconsistent) in [2 2] conditions, although the discrepancy is relatively small (overestimating p(low mean) by 0.054 or 31.37\%, and p(inconsistent) by 0.049 or 27.83\% in [2 2] condition).

As a control, we also applied posterior predictive checks on alternative models that consider only deterministic or only random noise, and these reduced models fail to capture all qualitative patterns (Supplementary Figure S13). Full details of this analysis can be found in Supplementary Materials section 3.2.

>>>>

% control the fontsize
\PassOptionsToPackage{textsize=scriptsize}{todonotes}
\usepackage[final]{changes}
\usepackage{changes}
\definechangesauthor[name={Siyu Wang}, color=red]{siyu}
\definechangesauthor[name={Bob}, color=blue]{bob}

@@

Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (see Supplementary Materials section 2.2). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. \added[id=siyu]{To ensure that the joint increase of random and deterministic sources of noise is not caused by a limitation of the fitting procedure, we also calculated the correlation between ground-truth values of random noise, and best-fitting values of deterministic noise (and vice versa), and they do not correlate (Supplementary Figure S9).}
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

@@

\subsubsection*{Model-based results} 
Posterior distributions over the group-level means of the deterministic and random noise standard deviation $\sigma_{det}$ and $\sigma_{ran}$ are shown in Figure \ref{fig:mb1} and Supplementary Figure S11. Consistent with our model-free results, we see that both random and deterministic noise are non-zero. Numerically, random noise is about 2-3 times larger than the deterministic noise. By computing the posterior distribution of $\sigma^2_{det}/(\sigma^2_{det}+\sigma^2_{ran})$, our data suggests that 14.25\% of the variability in random exploration is accounted for by deterministic noise ([4.90\%, 28.81\%], 95\% CI). In addition, we find that both random and deterministic noise increase with horizon. This increase was larger for random noise (mean = 7.13, 100\% of samples showed an increase in random noise with horizon) than deterministic noise (mean = 2.59, 98.64\% of samples showed an increase in deterministic noise with horizon). But intriguingly, the relative increase in both types of noise was similar (Figure \ref{fig:ratio}). That is, when we compute the relative increase in deterministic noise with horizon, $\sigma^{det}_{horizon6}/\sigma^{det}_{horizon1}$, it is very similar to the relative increase in random noise with horizon $\sigma^{ran}_{horizon6}/\sigma^{ran}_{horizon1}$. \added[id=siyu]{We also fit alternative models that fit random and deterministic noises separately for [1 3] and [2 2] conditions, and a model with an alternative definition of information bonus $dI$. These results are presented in the supplemental materials (Supplemental Figure. S17, S18). }

@@

To perform this `posterior predictive check', we created a set of simulated data by taking the subject-level parameters from the hierarchical Bayesian fits and having the model play the same sequence of games as seen by the subjects. We then applied the same model-free analysis as described in the previous sections to this simulated data set and compared the model's behavior to that of participants. As shown in Figure \ref{fig:mb3}, the model can account for all qualitative patterns in the data --- the increase in p(high info), p(low mean), and p(inconsistent) with horizon, and that p(inconsistent) is in between pure random and pure deterministic noise. \added[id=siyu]{The quantitative agreement is almost perfect for p(high info) and for p(inconsistent), but the model slightly overestimate p(low mean) in [2 2] conditions. This has to do with the skewness of the subject-level posterior distribution (see Supplemental Materials). }
%The quantitative agreement is almost perfect for p(high info) and for p(inconsistent) in the [1 3] condition, but the model seems to systematically overestimate p(low mean) and p(inconsistent) in [2 2] conditions, although the discrepancy is relatively small (overestimating p(low mean) by 0.054 or 31.37\%, and p(inconsistent) by 0.049 or 27.83\% in [2 2] condition).

As a control, we also applied posterior predictive checks on alternative models that consider only deterministic or only random noise, and these reduced models fail to capture all qualitative patterns (Supplementary Figure S14, S15). Full details of this analysis can be found in Supplementary Materials section 3.2.

### 2025-04-19 Plos Comp Bio - revision v1/supplemental/DetRanNoise_Supplementary.tex
u

@@

@@

@@

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=0.8\textwidth]{figures/RDBayes_2noise_6modelcomparison.jpg}
			\caption{
				Model comparison. A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
			\label{fig:posteriorcheck6models}
		\end{center}
	\end{figure}

@@

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.8\textwidth]{figures/RDBayes_2noise_hyperprior_6model_recovery.jpg}
			\caption{
				Our model qualitatively captures whether deterministic and random noise are present or not and whether either types of noise is dependent on horizon. A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
			\label{fig:6model}
		\end{center}
	\end{figure}
\end{document}

>>>>

	In general, increasing random noise leads to higher p(low mean) and lower p(consistent) between repeated games, and increasing deterministic noise leads to higher p(low mean) and higher p(consistent).

@@

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=0.65\textwidth]{figures/R1_shuffle_shuffled_recovery.jpg}
			\caption{Parameter recovery for shuffled data. To show that the joint increase of random and deterministic sources of noise is not caused by a limitation of the fitting procedure, we calculated the correlation between ground-truth values of random noise, and best-fitting values of deterministic noise (and vice versa). Ground-truth values are shuffled best-fit parameters. As expected, ground-truth random values do not correlate with recovered deterministic noises, showing that the increase of deterministic noise with horizon is genuine and not a by-product of increase of random noise, and vice versa.  }
			\label{fig:paramrecover_sub}
		\end{center}
	\end{figure}

@@

	In our model, the subject-level noises are assumed to follow a gamma distribution (to ensure positiveness), the posteriors are right skewed (maximal likelihood estimation or mode is smaller than the mean), Because of the skewness, simulating data with the “mean” is nosier than simulating with the true distribution. Simulating from the true distribution requires taking expected value over all possible noise values, for ease of implementation, we simulated data by taking random samples from the posterior distribution (instead of using the mean). The simulation was repeated 50 times and then averaged (Supplementary Figure \ref{fig:posteriorcheck6models}). Simulation using the "mode" or maximal likelihood estimation was also shown here (Supplementary Figure \ref{fig:posteriorcheck6models2}).

@@

	\begin{figure}[hp]
		\begin{center}
			\includegraphics[width=0.8\textwidth]{figures/R1_2noise_6modelcomparison_dist.jpg}
			\caption{
				Model comparison. A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
			\label{fig:posteriorcheck6models}
		\end{center}
	\end{figure}
	\begin{figure}[hp]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{figures/R1_2noise_6modelcomparison_map.jpg}
		\caption{
			Model comparison (using maximal likelihood estimation). A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
		\label{fig:posteriorcheck6models2}
	\end{center}
	\end{figure}

@@

	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.8\textwidth]{figures/RDBayes_2noise_hyperprior_6model_recovery.jpg}
			\caption{
				Our model qualitatively captures whether deterministic and random noise are present or not and whether either types of noise is dependent on horizon. A-D. both deterministic and random noise are horizon dependent, E-H. only random noise is horizon dependent, I-L. only deterministic noise is horizon dependent, M-P. neither random nor deterministic noise is horizon dependent, Q-T. only deterministic noise is assumed to be present, U-X. only random noise is assumed to be present.}
			\label{fig:6model}
		\end{center}
	\end{figure}

	\newpage
	\subsection{Alternative models: separate noises for [1 3] and [2 2] condition}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/R1_hyperprior.jpg}
			\caption{Model based analysis from a model that estimates random and deterministic noises separately for [1 3] and [2 2] conditions. The posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, E) and deterministic (C,G) noises are nonzero (A, C, E, G) and change with horizon (B, D, F, H).  However, random noise has both a greater magnitude overall (A, E) and a greater change with horizon (B, F) than deterministic noise. Both random and deterministic noises are higher in [1 3] compared to [2 2] conditions.}
			\label{fig:s8}
		\end{center}
	\end{figure}
	\newpage
	\subsection{Alternative models: use variance differences as dI}
	\begin{figure}[H]
		\begin{center}
			\includegraphics[width=0.7\textwidth]{figures/R1_hyperprior_dIvar.jpg}
			\caption{Model based analysis from a model that uses variance differences as dI. The posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.}
			\label{fig:s8}
		\end{center}
	\end{figure}

\end{document}
